---
title: "Appendix: Saving More Lives with Deceased Donor Kidney Transplantation"
author: "William F. Parker, MD, PhD,^1^, Yolanda Becker MD^1^, Robert D. Gibbons, PhD^1^" 
date: "^1^University of Chicago"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    #citation_package: natbib
  word_document:
    reference_docx: word_style_times_new_roman_1.docx
  html_notebook:
    theme: journal
    toc: yes
    toc_depth: 2
    toc_float: yes
bibliography: references.bib
csl: national-academy-science-letters.csl
urlcolor: blue
---

```{r packages, echo=FALSE, cache=FALSE, include=FALSE}
library(survival)
library(coxme)
library(tidyverse)
library(gtsummary)
library(knitr)
#library(furrr)
```

```{r global_options, include=FALSE, cache = FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, echo=FALSE, warning=FALSE, message=FALSE, linewidth=60)
options(scipen = 999)
```

```{r convenience_functions}
#Convenience functions 
numextract <- function(string){
  as.numeric(str_extract(string, "\\-*\\d+\\.*\\d*"))
}
comma <- function(x) format(x, digits = 3, big.mark = ",")
comma_1 <- function(x) format(x, digits = 1, big.mark = ",")
comma_2 <- function(x) format(x, digits = 2, big.mark = ",")
comma_p <- function(x){
  if (x < 0.001){
    return("< 0.001")
  }
  if (x<0.01 & x>=0.001){
    paste("=", format(x, digits = 3, big.mark = ","))
  }
  else{
    paste("=", format(x, digits = 2, big.mark = ","))
  }
} 
```


```{=tex}
\begin{center}
A commissioned paper for \textbf{A Fairer and More Equitable, Cost-Effective, and Transparent System of Donor Organ Procurement, Allocation, and Distribution}, an ad-hoc study committee of the National Academies of Sciences, Engineering, and Medicine.

July, 2021
\end{center}
```

```{=tex}
\setcounter{tocdepth}{2}
\tableofcontents
```

\pagebreak
# Data Preparation
```{r}
load("model_fit_extended.Rdata")

full_me_model <- me_models[[1]]
full_base_hz <- hz_list[[4]]
```

## Formal Description of the Data

With candidates by $i$, centers by $k$, and donors by $j$, denote the data as:

-   A vector of candidate covariates $X_{ik}(t)$, specifically the **EPTS variables** with their corresponding transformations

    -   $Max(Age\, (years)-25, 0)$
    -   $log(Years \, on \, dialysis + 1)$
    -   $I\{ Years , on , dialysis = 0\}$
    -   $Diabetes$
    -   $Prior \, Solid \, Organ \, Transplant$

-   Candidate death status ($Y_{ik} \in (0,1)$) and follow-up time ( $T_{ik}$). Includes death times after delisting for untransplanted candidates and post-transplant deaths

-   Time-dependent indicator variable for transplantation $Tx_{i}(t)$

    -   Donor quality $W_{j}$ (**KDRI Rao**)
    -   Ischemic time $I_{ikj}$, categorized as (\<12 hours, 12-22 hours, \>22 hours, or missing)
        
## Table S1: Sample of final analytic dataset

```{r example_data}
extended_to_model %>% 
  select(PX_ID, center, t_1, t_2, tx, dead, age, dialysis_time_years, diabetes, CAN_PREV_TX, tx_kdri, ischemic_time) %>%
  mutate(tx_kdri = ifelse(tx ==1, round(exp(tx_kdri), 2), ""),
         ischemic_time = ifelse(tx ==1, ischemic_time, ""),
         tx = case_when(
           tx == 1 & (t_2-t_1) <= 30 ~ "Transplant (day 0-30)",
           tx == 1 & (t_2-t_1) <= 180 ~ "Transplant (day 30-180)",
           tx == 1 ~ "Transplant (day 181+)",
           TRUE ~ "Waiting"
         )) %>%
  filter(PX_ID %in% c(500064,470987, 480003, 648174)) %>%
  select(PX_ID,center, age, dialysis_time_years, t_1, t_2, tx, tx_kdri, ischemic_time, dead)%>%
  kable(col.names = c("Patient", "Ctr.", "Current age", "Dialysis time (years)", "time start", "time stop", "Status (interval start)", "KDRI (linear)","Ischemic Time", "Dead (end interval)"), digits = 2)
```

Sample of four patient records in final analytic dataset.

-   Patient 470987 waited $684$ days before delisting and did not have a recorded death date so was censored at the time of delisting.

-   Patient 480003 waited $2,540$ days before transplant a KDRI Rao $1.59$ kidney with $12$ hours of ischemic time. Patient 480003 was still alive at last follow-up $4,001 - 2,540 = 1,461$ days post-transplant

-   Patient 500064 waited $206$ days for before DDKT with a KDRI Rao $1.1$ kidney with $14$ hours of ischemic time. After receiving this transplant, patient 500064 lived $2,476 - 206 = 2,270$ days before death.

-   Patient 648174 waited $117$ days before dying without a transplant


# Methodology

## Mixed-effect Cox Proportional Hazards Model

To estimate the primary outcome of survival benefit associated with DDKT, we fit a mixed-effects Cox proportional hazard model with time-dependent covariates and a non-proportional effect of transplantation [@parker2019].

$$
\begin{aligned} 
h_{ik}(t) = & h_0(t)*exp(\beta_{0k} + X_{ik}\beta + \\
& 1\{Transplant\}*(\beta_{1k} + \alpha_1X_{ik} + \Pi_2W_j + \zeta_1(W_j*X_{ij}) + \gamma I_{ijk}) +\\
& 1\{Day \; 0-30 \; post \: transplant\}*(\beta_{2} + \alpha_2 X_{ik} + \Pi_2 W_j + \zeta_2(W_j*X_{ij})) + \\
& 1\{Day \; 30-180\; post \: transplant\}*(\beta_{3} + \alpha_3 X_{ik} + \Pi_3 W_j + \zeta_3(W_j*X_{ij}))
\end{aligned}
$$

$$\beta_{0k} = \nu_{k0}$$ 
$$\beta_{1k} = \beta_1 + \nu_{k1}$$ 
$$(\nu_{k0}, \nu_{k1}) \sim N(0, \Sigma)$$
$$ \Sigma = \begin{pmatrix} \sigma_0^2 & \sigma_{01} \nonumber \\
\sigma_{01} & \sigma_1^2
\end{pmatrix}$$
this model includes:

-   A random intercept $\nu_{0k}$ for each center, representing risk of death without transplantation at that center. Note there is no fixed effect intercept (as this is a proportional hazards model, so the "intercept" is the baseline hazard function $h_0(t)$)

-   A random transplant effect $\beta_{1k}$, which represents the change in hazard from transplant at a particular center

-   A covariance matrix $\Sigma$ for the random effects, which allows for the intercept and transplant effect to be correlated

-   Effect of DDKT $\beta_{1k}$ modified by the following interaction effects: 

    -   Candidate factors $\alpha*X_{ij}$

    -   $\Pi$: KDRI $W_j$ and $\gamma$: ischemic time

    -    $\zeta$: candidate and donor interactions $W_j * X_{ik}(t)$


## Table S2: Mixed-effects cox proportional hazards model results

**Table S2A: Fixed Effect Coefficients**

```{r summarise_main_model}

coefs <- full_me_model$coefficients

w_break <- c(coefs, "")

names(w_break) <- c(names(coefs), "Transplant and interaction terms")

id  <- c( seq_along(coefs), 5+0.5 )
w_break <- w_break[order(id)]

w_break_se <- c(vcov(full_me_model) %>% diag() %>% sqrt(), "")

w_break_se <- w_break_se[order(id)]

tibble(variable = names(w_break),
       coef = w_break,
       se = w_break_se) %>%
  mutate(coef = as.numeric(w_break),
         se = as.numeric(se),
         up_ci = round(coef + 1.96*se, digits = 3),
         low_ci = round(coef - 1.96*se,digits = 3),
         coef = round(coef, digits =3),
         ci = paste0("(", low_ci, ",", up_ci, ")"),
       variable = case_when(
    variable == "age_floor_25" ~ "Max(age -25, 0)",
    variable == "log_dial_time" ~ "log(years on dialysis + 1)",
    variable == "no_dial" ~ "Never dialyzed (pre-emeptive listing)",
    variable == "CAN_PREV_TX" ~ "History of previous solid organ transplant",
    variable == "diabetes" ~ "Diabetes",
    variable == "age_floor_25:diabetes" ~ "Diabetes:Max(age -25, 0)",
    variable == "log_dial_time:diabetes" ~ "Diabetes:log(years on dialysis + 1)",
    variable == "no_dial:diabetes"~ "Diabetes:Never dialyzed (pre-emeptive listing)",
    variable == "diabetes:CAN_PREV_TX" ~ "Diabetes:History of previous solid organ transplant",
    variable == "tx" ~ "Transplantation",
    variable == "immediate_post_trans" ~ "Day 0-30 post-transplant",
    variable == "second_post_trans_int" ~ "Day 30-180 post-transplant",
    variable == "tx_kdri" ~ "KDRI linear component score",
    variable == "i_time_low" ~ "ischemic time < 12 hours",
    variable == "i_time_high" ~ "ischemic time > 22 hours",
    variable == "i_time_unknown" ~ "ischemic time not recorded",
    variable == "age_floor_25:tx" ~ "transplantation:Max(age -25, 0):",
    variable == "log_dial_time:tx" ~ "transplantation:log(years on dialysis + 1)",
    variable == "no_dial:tx" ~ "transplantation:Never dialyzed",
    variable == "diabetes:tx" ~ "transplantation:Diabetes",
    variable == "CAN_PREV_TX:tx" ~ "transplantation:Previous transplant",
    variable == "age_floor_25:diabetes:tx" ~ "Diabetes:Max(age -25, 0):Transplant",
    variable == "log_dial_time:diabetes:tx" ~ "Diabetes:log(years on dialysis + 1):Transplant",
    variable == "no_dial:diabetes:tx"~ "Diabetes:Never dialyzed:Transplant",
    variable == "diabetes:CAN_PREV_TX:tx" ~ "Diabetes:Previous transplant:Transplant",
    variable == "age_floor_25:tx_kdri" ~ "KDRI linear score:Max(age-25,0)",
    variable == "log_dial_time:tx_kdri" ~ "KDRI linear score:log(years on dialysis + 1)",
    variable == "no_dial:tx_kdri" ~ "KDRI linear score:Never dialyzed",
    variable == "diabetes:tx_kdri" ~"KDRI linear score:Diabetes",
    variable == "CAN_PREV_TX:tx_kdri" ~ "KDRI linear score:Previous transplant",
    variable == "age_floor_25:immediate_post_trans" ~ "Day 0-30 post-transplant:Max(age -25, 0):",
    variable == "log_dial_time:immediate_post_trans" ~ "Day 0-30 post-transplant:log(years on dialysis + 1)",
    variable == "no_dial:immediate_post_trans" ~ "Day 0-30 post-transplant:Never dialyzed",
    variable == "diabetes:immediate_post_trans" ~ "Day 0-30 post-transplant:Diabetes",
    variable == "CAN_PREV_TX:immediate_post_trans" ~ "Day 0-30 post-transplant:Previous transplant",
        variable == "age_floor_25:second_post_trans_int" ~ "Day 30-180 post-transplant:Max(age -25, 0):",
    variable == "log_dial_time:second_post_trans_int" ~ "Day 30-180 post-transplant:log(years on dialysis + 1)",
    variable == "no_dial:second_post_trans_int" ~ "Day 30-180 post-transplant:Never dialyzed",
    variable == "diabetes:second_post_trans_int" ~ "Day 30-180 post-transplant:Diabetes",
    variable == "CAN_PREV_TX:second_post_trans_int" ~ "Day 30-180 post-transplant:Previous transplant",
    variable == "tx_kdri:immediate_post_trans" ~ "KDRI linear score:Day 0-30 post-transplant",
    variable == "tx_kdri:second_post_trans_int" ~ "KDRI linear score:Day 30-180 post-transplant",
    TRUE ~ variable
  )) %>%
  select(Variable = variable, Coefficent = coef, "95% CI" = ci) %>%
  mutate_all(function(x) ifelse(is.na(x) == TRUE, "---", x)) %>%
  mutate(`95% CI` = ifelse(`95% CI` == "(NA,NA)", "---", `95% CI`)) %>%
  kable(digits = 2)
```

```{r extract_re}
intercept_Var <- VarCorr(full_me_model)$center[1,1]
transplant_Var <-VarCorr(full_me_model)$center[2,2]
correlation <- VarCorr(full_me_model)$center[1,2]

# tibble(name = c("Intercept (Waitlist) variance", "Transplant Effect", "Correlation"),
#        values = c(VarCorr(full_me_model)$center[1,1], VarCorr(full_me_model)$center[2,2], VarCorr(full_me_model)$center[1,2]))
```

**Table S2B: Center-level random effects**

| Term                                    | Value                                   |
|-----------------------------------------|-----------------------------------------|
| Variance of waitlist risk ($\nu_0$ )    | `r intercept_Var %>% round(digits =2)`  |
| Variance of transplant effect ($\nu_1$) | `r transplant_Var %>% round(digits =2)` |
| Correlation ($Corr(\nu_0, \nu_1)$ )     | `r correlation %>% round(digits =2)`    |

```{r c_statistic}
full_me_model <- me_models[[1]]
lp_coxme <- predict(full_me_model)

for_cstat <- extended_to_model %>%
  select(dead, t_1, t_2, age_floor_25, log_dial_time, no_dial, diabetes, CAN_PREV_TX, tx, tx_kdri,
         i_time_low, i_time_high, i_time_unknown, center,
         immediate_post_trans, second_post_trans_int) %>%
  na.omit() %>%
  filter(t_2 > t_1) %>%
  select(dead, t_1, t_2) %>%
  cbind(lp_coxme)

cstat_full <- concordance(Surv(t_1, t_2, dead) ~ lp_coxme, data = for_cstat, reverse = TRUE)
```

This fit is from N = `r full_me_model$n[2] %>% comma()` adult candidates listed for kidney alone deceased donor transplant from 2005 to 2010, there were N = `r comma(full_me_model$n[1])`. The discriminatory accuracy of the model as measured by Harrel's C-statistic was `r round(cstat_full$concordance, digits =2)`. The between-center variance in survival benefit of transplant on the log hazard scale ($B_i = \nu_{1i} - \nu_{0i}$) is the variance of the difference between the two center effects, or

$Var(\nu_{i1} - \nu_{i0}) = Var(v_{1i}) + Var(\nu_{0i}) - 2*Corr(v_{1i}, v_{0i})*\sigma_{v_{0i}}\sigma_{v_{1i}}$

```{r var_tx_benefit}
variance_survival_benefit <- function(me_cox){
  variance_est <- me_cox$vcoef[[1]]
  
  var_wait <- variance_est[[1,1]]
  var_tx <- variance_est[[2,2]]
  cov_tx_wait <- variance_est[[1,2]]*sqrt(variance_est[[1,1]])*sqrt(variance_est[[2,2]])
  
  return(var_wait + var_tx - 2*cov_tx_wait)
}


```

For the model estimated here, the between-center standard deviation of the survival benefit on the log hazard scale is `r variance_survival_benefit(full_me_model) %>% sqrt() %>% round(digits = 2)`


# Lives Saved by Transplant within 5 years (LiST-5) calculation 

While a hazard ratio of transplantation for each specific candidate-donor pair can be calculated directly from the coefficients in **Table S2**, an estimate of the baseline hazard function $\hat{h}_0(t)$ is required to calculate the improvement in absolute survival with DDKT. Following the previously published methodology of @parker2019, we construct estimates of survival with transplant $S(t|transplant)_{ijk}$ and survival without transplant $S(t|waitlist)_{ijk}$ from the estimated model coefficients and a Nelson-aalen estimate of the baseline hazard function $\hat{h}_{0}(t)$ (**Figure S1**).

Specifically, assume candidate $X_{ik}$ has waited $t_{tx}$ days and receives an offer for kidney $W_j$ which will suffer ischemic time $I_{ikj}$ in transit. The model can generate counterfactuals for survival with and without transplant.

1.  Calculate the hazard function for the patient with and without transplant

    -   **with transplant**: $$\hat{h}_{ijk}(t|transplant) = \hat{h}_0(t)*(exp(\hat{\beta}_{0k} + X_{ikt}\hat{\beta} + \hat{\beta}_{1k}  + \hat{\alpha}(X_{ik})) \hat{\Pi}*(W_{j}) + \hat{\gamma}(I_{ikj})) + \hat{\zeta}(W_{j}*X_{ik}))$$

    -   **without transplant** 
    $$\hat{h}_{ijk}(t|waitlist) = \hat{h}_0(t)*(exp(\hat{\beta}_{0k} + X_{ikt}\hat{\beta} )$$


2.  Construct survival functions from the hazard functions, comparing transplantation to waiting without transplantation.

    -   **with transplant**: $$\hat{S}(t|transplant)_{ijk} = exp(-\int_{t_{tx}}^t \hat{h}_{ijk}(t|transplant)) dt$$

    -   **without transplant**: $$\hat{S}(t|wait)_{ik} = exp(-\int_{t_{tx}}^t \hat{h}_{ik}(t|wait)) dt$$

Note that because the effect of transplantation is non-proportional, the hazard ratio varies inside the integral, i.e. if $t > t_{tx} + 180$

$$
\begin{aligned} 
\hat{S}(t|transplant)_{ijk} = exp(-(&\int_{t_{tx}}^{t_{tx}+30} \hat{h}_{ijk}(t|Day \; 0-30 \; post \: transplant) +\\ 
& \int_{t_{tx}+30}^{t_{tx}+180} \hat{h}_{ijk}(t|Day \; 30-180 \; post \: transplant) + \\
& \int_{t_{tx} + 180}^t \hat{h}_{ijk}(t|Day >180 \:post \: transplant)) \:) dt
\end{aligned}
$$

3.  Calculate the estimated absolute survival benefit at the specified points in time

    -   $LiST-1_{ijk} = \hat{S}(365 \, days|transplant)_{ijk} - \hat{S}(365 \, days|wait)_{ik}$
    -   $LiST-5_{ijk} = \hat{S}(1825 \,days|transplant)_{ijk} - \hat{S}(1825 \,days|wait)_{ik}$

\pagebreak
### Figure S1: Non-parametric estimated of the baseline survival function $\hat{S}_0(t)$

```{r baseline_hazard_plot}
full_base_hz %>%
  mutate(survival = exp(-hazard),
         time = time/365) %>%
  ggplot(aes(x = time, y = survival)) +
  geom_step() + lims(y = c(0,1), x = c(0,5)) + 
  labs(x = "Years after listing") + 
  ggthemes::theme_gdocs() 
```
Baseline survival function for the model estimated using Cox's extension of the Nelson-Aalen estimator. This represents the survival for the base case deceased donor kidney transplantation candidate (< 25 years old, just started dialysis, without diabetes) in the absence of transplantation during the first five years on the waitlist

\pagebreak
```{r log_hazard_function}
sum_hazard <- function(param_list,start_hazard, covariate_values, model_coef){
        h_out <- start_hazard
        
        for (wp in param_list ) {

          params <- str_split(wp, ":")
          value <- prod(sapply(params[[1]], function(x) unname(covariate_values[x])))
          #print(value)
          beta <- unname(model_coef[wp])
          #print(beta)
          add_hz <- value*beta
          #print(add_hz)
          h_out <- h_out + add_hz
  
        }
        return(h_out)
}


log_hazard <- function(ctr_id, Age, d_time, diabetes, CAN_PREV_TX, 
                            ischemic_time, kdpi,
                            period = "wait",
                            model= full_me_model,
                            icuts = i_time_cuts,
                            ...){
  
  no_dial <- ifelse(d_time == 0, 1,0)
  age_floor_25 <- ifelse(Age < 25, 0, Age - 25)
  log_dial_time <- log(d_time + 1)
  tx_kdri <- KDRI_raw(kdpi)

  i_time_low <- ifelse(ischemic_time <= icuts[[1]], 1, 0)
  i_time_high <- ifelse(ischemic_time > icuts[[2]], 1, 0)
  
  user_patient_params <- c(age_floor_25, log_dial_time, no_dial, diabetes, CAN_PREV_TX, 
                           i_time_low, i_time_high, tx_kdri)
  
  names(user_patient_params) <- c("age_floor_25", "log_dial_time", "no_dial", "diabetes", "CAN_PREV_TX",
                                  "i_time_low", "i_time_high", "tx_kdri")
  
  user_patient_params[["i_time_unknown"]] <- 0

  fixed_effects <- fixef(model)
  x_vars <- names(fixed_effects)
  ref <- data.frame(ranef(model)[[1]])
  ref$center <- row.names(ref)
  
  if (ctr_id == "mean center"){
    tx_ref <- 0
    int_ref <- 0
  } else {
      tx_ref <- filter(ref, center == ctr_id)$tx
      int_ref <- filter(ref, center == ctr_id)$Intercept
  }
  
  if (period == "wait"){
    entered_params <- c(user_patient_params, "tx" =0, "immediate_post_trans"=0, "second_post_trans_int" =0)
    entered_params[["tx_kdri"]] <- 0
    entered_params[["i_time_low"]] <- 0
    entered_params[["i_time_high"]] <- 0
    st_hazard <- int_ref
  }
  
  if (period == "post_1"){
      entered_params <- c(user_patient_params, "tx" =1, "immediate_post_trans"=1, "second_post_trans_int" =0)
      st_hazard <- int_ref + tx_ref
  }
  
  if (period == "post_2"){
      entered_params <- c(user_patient_params, "tx" =1, "immediate_post_trans"=0, "second_post_trans_int" =1)
      st_hazard <- int_ref + tx_ref
  }
  
    if (period == "post_3"){
      entered_params <- c(user_patient_params, "tx" =1, "immediate_post_trans"=0, "second_post_trans_int" =0)
      st_hazard <- int_ref + tx_ref
  }
  # print(x_vars)
  # print(st_hazard)
  # print(entered_params)
  # print(fixed_effects)
  h_out <- sum_hazard(x_vars, st_hazard, entered_params, fixed_effects)
  return(h_out)

}
```

```{r survival_benefit_function}
percent <- function(x, digits = 1, format = "f", ...) {      
  paste0(formatC(x * 100, format = format, digits = digits, ...), "%")
}

survival_benefit <-  function(t_transplant, 
                              ctr_id, 
                              Age, d_time, diabetes, CAN_PREV_TX, 
                              ischemic_time, kdpi,
                              output = "plot",  
                              me_model = full_me_model, 
                              isch_cuts = i_time_cuts,
                              b_hazard = full_base_hz,
                              post_tx_int_1 = 30,
                              post_tx_int_2 = 180){
  
  log_hr_w <- log_hazard(ctr_id, Age, d_time, diabetes, CAN_PREV_TX, ischemic_time, kdpi, 
                                     model = me_model, icuts = isch_cuts, period = "wait")
  
  log_hr_post_1 <- log_hazard(ctr_id, Age, d_time, diabetes, CAN_PREV_TX, ischemic_time, kdpi, 
                                     model = me_model, icuts = isch_cuts, period = "post_1") 
  
  log_hr_post_2 <- log_hazard(ctr_id, Age, d_time, diabetes, CAN_PREV_TX, ischemic_time, kdpi, 
                                     model = me_model, icuts = isch_cuts, period = "post_2")              

  log_hr_post_3 <- log_hazard(ctr_id, Age, d_time, diabetes, CAN_PREV_TX, ischemic_time, kdpi, 
                                     model = me_model, icuts = isch_cuts, period = "post_3")    
                              
  cum_hz <- filter(b_hazard, time == t_transplant)$hazard
  
  df <- b_hazard %>%
  mutate(hazard = hazard- lag(hazard),
                   hazard = ifelse(is.na(hazard), 0, hazard)) %>%
  filter(time >= t_transplant) %>%
  mutate(
    time = time - t_transplant,
    hazard_wait = hazard*exp(log_hr_w),
    hazard_tx = case_when(
      time < post_tx_int_1 ~ hazard*exp(log_hr_post_1),
      time < post_tx_int_2 ~ hazard*exp(log_hr_post_2),
      TRUE ~ hazard*exp(log_hr_post_3)
      ),
    H_tx = cumsum(hazard_tx),
    H_wait = cumsum(hazard_wait),
    S_tx = exp(-H_tx),
    S_wait = exp(-H_wait),
    S_tx_better = ifelse(S_tx >= S_wait, S_tx, NA),
    S_tx_worse = ifelse(S_tx <= S_wait, S_tx, NA),
    S_wait_better = ifelse(S_wait >= S_tx, S_wait, NA),
    S_wait_worse = ifelse(S_wait <= S_tx, S_wait, NA))
  
  if (output == "dataframe"){
    return(df)
  }
  
  if (is.numeric(output)){
    last_ob <- df %>%
      filter(time == output)
  
    if (nrow(last_ob) > 0){
      return(last_ob %>% 
               select(S_tx, S_wait) %>%
               mutate(benefit = S_tx - S_wait))
    } else {
      
      return(tibble(S_tx = NA, S_wait = NA, benefit = NA) %>%
               mutate_all(as.numeric))
    }
  }
  

  # if (output == "wait_5"){
  #       last_ob <- df %>%
  #     filter(time == 1825)
  #     
  #       if (nrow(last_ob) > 0){
  #           return(last_ob$S_wait)
  #         } else {
  #           return(NA)
  #         }
  # }
  
  plot <- df %>%
              filter(time < 365*5) %>%
              mutate(years_from_transplant = time/365,
                     tx_dot = ifelse(years_from_transplant==0, 1, NA),
                     benefit = S_tx - S_wait) %>%
              ggplot(aes(x = years_from_transplant)) +
              geom_ribbon(aes(ymin = S_wait_worse, ymax = S_tx_better, fill = "Survival Benefit"), alpha = 0.5) +
              geom_step(aes(y = S_tx, color = "Post-Transplant"), size = 1.1) +
              geom_step(aes(y = S_wait, color = "Waitlist"), size = 1.1) +
              geom_point(aes(y= tx_dot, shape= "Transplant"), size = 5, colour = "darkgreen") +
                scale_x_continuous(breaks = seq(0, 5)) + 
               ggthemes::theme_gdocs() +
                   labs(y = "Survival", x = "Time (years after transplant)", shape = "") +
                guides(shape = guide_legend(order = 1), color = guide_legend(order = 2), fill = guide_legend(order = 3))
  
  if (output == "plot"){
      return(plot +
               scale_color_manual(name = "Survival function", values=c("blue", "red")) +
               scale_fill_manual(name = NULL, values=c("skyblue")) +
               lims(y = c(0,1))
             )
  }
  
  if (output == "plot_w_benefit"){
      return(plot + 
               geom_step(aes(y = benefit, color = "Survival benefit"), size = 1.1) + 
               scale_color_manual(name = "Survival function", values=c("blue", "darkgreen", "red")) +
               scale_fill_manual(name = NULL, values=c("darkgreen")) 
             )
  }

  if (output == "rmst"){
    rmst_df <-df %>%
        mutate(days = (S_tx - S_wait)*(time-lag(time)),
               days = ifelse(row_number() == 1, 0, days),
         rmst = cumsum(days))
      return(
         paste0("DDKT would improve survival by ", round(filter(rmst_df, time == 5*365)$rmst/365, digits = 1), " years on average within the first 5 years following transplant")
        )
    }
  if (output == "table"){
        benefit_df <- df %>%
      filter(time %in% c( 365, 3*365, 5*365))


    return(
      benefit_df %>%
      mutate(S_tx = S_tx,
             S_wait = S_wait,
             surv_benefit = S_tx - S_wait,
             years_from_transplant = time/365) %>%
      select("Years from transplant" = years_from_transplant,
             "Survival (transplant)" = S_tx,
             "Survival (waitlist)" = S_wait,
             "Absolute Survival Benefit" = surv_benefit) %>%
          mutate(across(c(2,3,4), percent)) %>%
        kable()
      )

  }

}
```

\pagebreak
## Figure S2: The non-proportional effect of transplantation 

```{r}
load("EPTS_values.Rdata")
extended_to_model <- extended_to_model %>%
  cbind(EPTS) 

recips <- extended_to_model %>% 
  filter(tx ==1 & immediate_post_trans ==1) %>%
  select(PX_ID, center, age, age_floor_25, dialysis_time_years, log_dial_time, no_dial, diabetes, CAN_PREV_TX, 
         tx,immediate_post_trans, second_post_trans_int, 
         ischemic_time, i_time_low, i_time_high, i_time_unknown, tx_kdri, 
         t_1, t_2, dead, 
         raw_epts, EPTS,
         REC_TX_DT) %>%
  na.omit()
i_time_cuts <- quantile(recips$ischemic_time, na.rm = TRUE, probs = c(0.25, 0.75))
```


```{r}
survival_benefit(700, "mean center", 55, 3, 0, 0, 20, 40) + 
  lims(y = c(0.95, 1), x = c(0, 250/365))
```
Visualization of the non-proportional hazard of transplantation for a 55-year old recipient with 3 years of dialysis time transplanted after 700 days of waiting. In the first 30 days post-transplant, the risk of death is actually higher compared to remaining on the wait-list. In days 30-180, the benefit of transplantation outweighs the post-surgical risks. After day 180 post-op, the benefit of transplantation increases further.
\pagebreak


# Shared decision making tool: Take the offer or wait for a better one?

## Formal description

When a candidate receives a deceased donor kidney offer, they have two options:

1.  Accept kidney $W_{j}$ after $t_a$ time on the wait-list. This generates the post-transplant survival function and survival benefit generated by our standard model.

$$\hat{S}(t|accept \, W_j)_{ijk} = exp(-\int_{t_{a}}^t \hat{h}_{ijk}(t|transplant \, W_{j})) dt$$

2.  Reject kidney $W_{j}$ and wait until time $t_b = t_a + \Delta t$ to get a better kidney $W_{m}$. This patient experiences a several discontinuities in estimated hazard, starting off with waitlist risk during the waiting period ( $\hat{h}(wait)$ for time $(t_a, t_b)$ )and then transitioning to post-transplant risk after accepting the better kidney $W_m$ ($\hat{h}(transplant \, W_m)$ for all time after $t_b$)

$$\hat{S}(t|wait \,for \, W_m)_{ijk} = exp(-(\int_{t_{a}}^{t_b} \hat{h}_{ijk}(t|wait)dt + \int_{t_b}^t\hat{h}_{ijk}(t|transplant \, W_{m})dt))$$

\pagebreak
# References
